---
title: "Boston Housing Prices Regression"
format: html
editor: visual
---

## Predicting housing prices with a regression

#### The data

Suppose we want to predict the price of housing in Boston in the mid to late 1970s. Luckily, we have a rather small dataset of housing prices (506 samples) from the mid-70s that we can use for a regression.

-   The samples are broken down into 404 training samples and 102 test samples

-   Each feature has a different scale and will need to be regularized.

```{r}
# Load the data
library(keras3)
boston <- dataset_boston_housing()
c(c(train_data, train_targets), c(test_data, test_targets)) %<-% boston
```

Inspect the data

```{r}
str(train_data)
str(test_data)
str(train_targets)
```

-   There are 13 numerical features associated with each sample

-   The targets are the median value of the homes in thousands of dollars

    -   Note: The prices seem super low because they are from the mid-70s and haven't been adjusted for inflation.

#### Preparing the data
